\chapter{Implicit Computational Complexity}\label{chap:icc}
Implicit computational complexity (ICC) studies how to guarantee resource bounds
without appealing to external machine models.
Instead of analysing running time or space after the fact, ICC designs
languages and recursion schemes whose syntactic constraints ensure that every
definable function belongs to a chosen complexity class.
The aim is a principled foundation for programming languages that ``build in''
complexity guarantees by construction.

Our goals in this chapter are twofold.
For one, we want to introduce
the most promising ICC approaches we have found towards obtaining a language for
\complexity{FL} or for \complexity{FP}. Indeed, the characterizations we will study
resulted in the creation of two actual programming languages. They are purely of academic
interest, but there are close to none alternative languages for complexity classes
that reached the stage of being actually implemented.
In the rest of this chapter we explain why the approach we studied most ---
of using techniques from ICC --- did not lead to a practical programming language for our purposes,
even though it initially seems very appealing.
However, more importantly, we want to share \emph{negative} evidence of
usefullness of the characterizations studied in ICC towards our ultimate goal
of not just creating a programming language, but using it to \emph{certify}
the complexity of known algorithms.

We will begin by sharing our experience with ICC,
then proceed to discuss the characterizations we have studied, most importantly:
in~\autoref{def:bc-eps} we examine a programming language based on the first
function algebra (untyped) for \complexity{FL} due to Neergaard; in~\autoref{sec:intml} we study
Dal Lago and Schöpp's programming language \texttt{IntML} with the type system 
guaranteeing the capturing of \complexity{FL} and \complexity{FNL} (nondeterministic version of \complexity{FL},
which we don't introduce formally).

\begin{remark}
  Recall that, given a description of a Turing machine, the problem of deciding whether it halts
  is undecidable. Hence, deciding whether an arbitrary computer program belongs to the complexity
  class \compP{} is also undecidable. This is only a problem, however, if we take \emph{arbitrary}
  programs as input. If we limit the scope to programs written in a special, limited programming
  language, we can easily design the language so that it does not admit constructs such as
  \texttt{while}-loops or general recursion, and therefore every program in it necessarily terminates.
  Moreover, membership in the syntax of such a language can be easy to decide. In this way, the
  undecidability problem is not resolved, but ``shifted'' to the difficulty of programming: given a
  \texttt{C} program implementing an algorithm, it becomes undecidable whether there exists a
  corresponding program in our restricted language.
\end{remark}


That was the approach that took us the most time to study and did not result in obtaining
``practical programming languages''. We want to place ourselves in the program of ``we need a better way of
communicating negative results (failed research)'', and signal why, despite the apparent
usefulness of the results for our purpose, our goal failed. The reason is the difference between
\emph{intenstional} and \emph{extensional} expressive power of characterizations of complexity classes.
With type systems capturing low complexity classes, it becomes very apparent that these function
classes are only \emph{extensionally} equivalent to their corresponding complexity classes.
For an example of how easily we can lose \emph{intensional} expressive power in a language with
linear types, please look at~\autoref{lst:haskell-linear-example}.
They shift the problem of undecidability of deciding if a given Turing machine works in some complexity,
to the problem if the algorithm implemented by it is transferrable to the particular programming language.

\subsection{Extensional and intensional equality of sets of functions}\label{subsec:intensional}

The field appear very fragmented and the approaches don't scale:~\cite{DalLagoMartini2006MuTutorial}.

\begin{remark}
    We don't introduce tiered recursion, ramified recursion and Leivant's works in general here:~\cite{LeivantRamyaa11}.
    See also~\cite{3ffa7833-e2d2-3419-abb5-7f266190ba48} for discussion on tiering as recursion technique.
\end{remark}



Accessible introductions to ICC include the three-part presentation~\cite{martini2006implicit1,martini2006implicit2,martini2006implicit3},
the talk~\cite{ronchi2019logic},
and a short overview~\cite{DalLago2012}.


\begin{remark}
    We focused on the characterizations of \complexity{FL} and \complexity{FP}.
\end{remark}

% \subsection{Related works on specifically the \complexity{FL} class}
% Early function algebras for \complexity{FL} appeared in~\cite{10.1145/1008293.1008295} and~\cite{lind1974logspace},
% but these were explicit characterizations.

% In~\cite{10.1007/978-3-662-46678-0_27}, an interesting approach using coinduction is utilized to capture \complexity{FL}.

% In~\cite{hofmann2006logspace} a good overview of languages for \complexity{FL} is presented,
% and in~\cite{schoepp2006spaceefficiency}, the history of \complexity{FL} characterizations is traced.

Description of logspace, ptime (decisive):~\cite{Jones99}; logspace, linspace:~\cite{Kristiansen05}.~\cite{Bonfante06}
For a broad literature survey, see~\cite{bloch1994function}.




\subsection{Characterizations not easily adjustable for a programming language}
Before the seminal works that founded the field of Implicit Complexity, many characterizations
of complexity classes had been known already. All of them suffered at least one of the two problems:
either it only characterized a class of relations in a given complexity (as opposed to functions),
or the characterization wasn't purely syntactic. We will refer to the latter of being ``explicit''
instead of ``implicit''.

\subsection{Characterizations of classes of relations}
Characterizations of classes of relations, such as \complexity{P} (as opposed to \complexity{FP}),
are not of interest to us because they don't generalize at all to a programming language
allowing to write functions with output. Nevertheless, we investigated the concepts used
there and describe some of them briefly in this subsection.

Polynomial-time relations have been characterized without explicit
size bounds in~\cite{doi:10.1137/0216051}.
In~\cite{COMPTON1990241}, uniform \(\complexityi{NC}{1}\) was characterized,
and in~\cite{ALLEN19911} uniform \(\complexity{NC}\), though their definitions
still concealed polynomial bounds and targeted relations instead of functions.

In more modern works, decisive complexity classes have been successfully characterized in~\cite{JONES1999151} by a fragment of Lisp in~\(\complexity{L}\) and
\(\complexity{P}\). The same concept has been extended to account for
for nondeterminism in~\cite{10.1007/11784180_8}.
The authors of~\cite{kristiansenvoda2005} investigated both imperative
and functional programming languages whose fragments yield hierarchies containing \emph{decisional} \complexity{L},
\complexity{LINSPACE}, \complexity{P}, and \complexity{PSPACE}.
Related contributions include~\cite{kristiansen2005neat} and~\cite{Oitavem+2010+355+362}.



The modern study of ICC begins with two breakthroughs:~\cite{151625}~and~\cite{10.1007/BF01201998}
gave the first implicit characterisations of polynomial-time computable functions.

However, the idea of Bellantoni and Cook seemed to best align with being the foundation
of a practical programming language. Hence, we decided to solely focus on it and its successors.

Since then (since BC, leivant), numerous classes have been captured implicitly; see, for
example,~\cite{NIGGL201047}~and~\cite{10.1016/j.ic.2015.12.009}
for overviews of \(\complexity{FP}\) and \(\complexity{FNC}\) characterisations.

See~\cite{10.1007/s00153-022-00828-4} for implicit characterizations of counting classes such as $\complexity{\mathbin{\#}P}$ (not introduced here).

\input{chapters/icc-recursion-theory.tex}
\input{chapters/icc-linear-logic.tex}