\chapter{Preliminaries}\label{chap:preliminaries}
In this chapter we introduce the standard definitions for logic and complexity theory, necessary to
fix the language we will use in the rest of the work. 
We restrict attention to the logical frameworks needed later (single-sorted first-order logic and
its two-sorted counterpart) and omit the analogous presentations for propositional calculus or
full second-order systems. The semantics is entirely classical
(as in: not intuitionistic). 

In~\autoref{sec:defs-preliminaries-circuits} we first introduce the so-called \emph{non-uniform}
circuit families (e.g.\ \FACipoly), then define a weak notion of uniformity and the appropriate
\emph{uniform} circuit families (e.g.\ \FACi).
The notion of uniformity used in this thesis (cf.~\autoref{def:logspace-uniformity})
and the results using it are usually not interesting. However, in this work
we will not focus too much on that problem. We leave out the complicated
divagations about the appropriate (i.e.\ much weaker) notions of uniformity to~\autoref{chap:uniformity}.

\begin{remark}[Bibliography]
The definitions in~\autoref{sec:defs-single-sorted} are in the style of~\cites[Section~2B]{CookNguyenDraft}[Section~II.2]{Cook_Nguyen_2010}.
The definitions in~\autoref{sec:defs-two-sorted} are in the style of~\cites[Section~4B]{CookNguyenDraft}[Section~IV.2]{Cook_Nguyen_2010}.
We will mostly need them for~\autoref{chap:bounded-arithmetic}, but also for~\autoref{chap:descriptive-complexity} --- 
the results discussed in the latter are mostly from~\cite{Immerman1999-IMMDC},
where different style of definitions is used.
However,~\cite{Immerman1999-IMMDC} doesn't introduce two-sorted logic.
The single-sorted definitions differences are mostly negligible and the consistent treatment of
single- and two-sorted logic, is important for~\autoref{chap:bounded-arithmetic}.

Much effort has been put into
considering different versions of the definitions in~\autoref{sec:preliminaries-turing}
to keep them consistent
with the definitions of functional complexity classes studied later in~\autoref{sec:functional-complexity-classes},
e.g.\ the class \complexity{FNP} defined in~\autoref{def:complexity-fnp}.
These are much less standard, and often have no clear consensus in the literature.

The definitions of decisional circuit classes are from~\cite{10.5555/520668}.
\end{remark}


\input{chapters/preliminaries-single-sorted-logic.tex}
\input{chapters/preliminaries-two-sorted-logic.tex}
\input{chapters/preliminaries-turing-complexity.tex}
\input{chapters/preliminaries-circuits.tex}
